### 2022-03-21 paper/localpsf.tex
removing old stuff

% REQUIRED
\begin{abstract}
	We present a fast matrix-free method for computing hierarchical matrix (H-matrix) approximations of operators that have locally supported non-negative integral kernels. Such operators arise, for example, as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained inverse problems. 
%	H-matrix approximations are highly desirable because 
	Fast H-matrix methods can add, multiply, invert, and factorize H-matrices, compute matrix-vector products, solve linear systems, and perform other useful matrix computations. 
%	The computational cost of these operations is $O(N \log^a N)$, where the matrix is $N\times N$, and $a \in \{1,2,3\}$. 
%	Classical H-matrix construction methods 
%	require evaluating the integral kernel $O(N \log N)$ times, but this is not feasible 
%	are infeasible for large-scale problems in which the operator is only accessible via matrix-vector products (``matrix-free''), because these methods require evaluating the integral kernel of the operator at a large number of points. We overcome this problem by c
	Our method computes impulse responses of the operator at a collection of scattered points, then interpolates these known impulse responses to evaluate entries of the integral kernel at other points where the impulse response is not known. 
%	We improve upon classical product-convolution approximations by translating with respect the estimated mean of the impulse response, rather than the location of the impulse response point source. 
	Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. 
%	We construct a-priori ellipsoid estimates of the supports of the impulse responses (i.e., support ellipsoids are estimated \emph{before} any impulse responses have been computed). 
	By solving an ellipsoid packing problem, we choose as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. 
	We apply the method to approximate Hessians in large-scale PDE-constrained inverse problems with highly informative data. Numerical results demonstrate that our method substantially outperforms existing state-of-the-art Hessian approximation methods which are based on low-rank approximation. We are able to form high quality approximations of high rank operators using only a small number of operator actions products.
\end{abstract}

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$) and let $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the following form:
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, $\Aop u \in L^2(\Omega)'$ is the linear functional that results from applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$. We seek to approximate operators $\Aop$ that are accessible \emph{matrix-free}. That is,
%access to $\Aop$ is available through ``operator actions.'' That is,
we have a black-box computational procedure through which we may evaluate the maps
\begin{equation*}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv
\end{equation*} 
for arbitrary functions $u$ and $v$, but we cannot easily compute kernel entries $\Aker(y,x)$. In applications, evaluating these maps may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representation of $\Aop$, but cannot easily compute individual matrix entries.
%After discretization (see Appendix \ref{app:discretized_operations}), $\Aop$ becomes a dense matrix, $\bm{A}$, that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means that one can perform matrix-vector products of $\bm{A}$ and $\bm{A}^T$ with arbitrary vectors, but one cannot easily\footnote{One could compute $\bm{v} = \bm{A}\bm{e}_j$, where $\bm{e}_j=(0,\dots,0,1,0,\dots,0)$ is the unit vector which has $j^\text{th}$ component equal to one and all other components equal to zero, then extract $\bm{A}_{ij} = \bm{v}_i$. However, this process is wasteful because one computes the entire vector $\bm{v}$, then discards all but one component.} access matrix entries $\bm{A}_{ij}$.

%For example, in reduced space approaches to PDE-constrained optimization and inverse problems, Hessian information is available only through the application of the Hessian to vectors, and one such Hessian application requires solving two PDEs \cite{HESSIANADJOINT}.

Our method, described in Section \ref{sec:method}, belongs to a class of methods that approximate an operator by interpolating impulse responses of the operator associated with a scattered collection of points. The \emph{impulse response} of $\Aop$ associated with the point $x$ is the function $\impulseresponse_x:\Omega \rightarrow \mathbb{R}$, 
\begin{equation}
\label{eq:impulse_response_defn}
\impulseresponse_x(y) := \Aker(y, x),
\end{equation} 
which results from keeping $x$ fixed and viewing $\Phi(y,x)$ as a function of $y$.
%for a collection of points $x_i$ scattered throughout $\Omega$.
%Further, let $\impulseresponse_x:\Omega \rightarrow \mathbb{R}$,
%\begin{equation}
%\label{eq:impulse_response_defn}
%\impulseresponse_x(y) := \Aker(y, x),
%\end{equation} 
%be the \emph{impulse response} of $\Aop$ at $x$. 
The function $\impulseresponse_x$ is called the impulse response because straightforward analysis shows that it may be expressed as follows:
\begin{equation}
\label{eq:impulse_response_delta_action}
\impulseresponse_x = \left( \Aop \delta_x^* \right)^*,
\end{equation}
where $\delta_x$ is the the delta distribution (i.e., point source, impulse) centered at $x$, $\Aop \delta_x^* \in L^2(\Omega)'$ is an abuse of notation that denotes the result of applying $\Aop$ to the distribution $\delta_x$, and $\left( \Aop \delta_x^* \right)^* \in L^2(\Omega)$ denotes the Riesz representation of $\Aop \delta_x^*$ with respect to the $L^2$ inner product.\footnote{Recall that the Riesz representative of a functional $\sigma \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\sigma^* \in L^2(\Omega)$ such that $\sigma(v) = \left(\sigma^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} Note that while the domain of $\Aop$ is defined as $L^2(\Omega)$, which does not contain $\delta_x$, if $\Aker$ is sufficiently regular then the action of $\Aop$ may be extended to distributions. This is described in Appendix \ref{app:distributions}.

@@

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch1.png}
		%		\caption{Non-symmetric}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch12.png}
		%		\caption{Non-symmetric}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch19.png}
		%		\caption{Non-symmetric}
	\end{subfigure}
	\caption{Batches, $\eta_b$, of normalized impulse responses, $\phi_{x_i}/V(x_i)$, for the Stokes inverse problem Hessian. Left: 2nd batch. Mid: 13th batch. Right: 19th batch. Brightness scale differs between the subfigures. Black stars indicate the sample point locations, $x_i$. Dashed gray ellipses indicate the support ellipsoids, $E_{x_i}$. The large circle is the boundary of the domain $\Omega$. Early batches tend to have lots of ellipsoids near the boundary, as compared to later batches. This is because our greedy ellipsoid packing algorithm tries to pick sample points that are far away from sample points in previously chosen batches.}
	\label{fig:batches_intro}
\end{figure}

@@

%
% Because we build a H-matrix, and 
%
%In this paper we form hierarchical matrix approximations, $\mathbf{A}_H$, of operators $\Aop$ that satisfy the following three properties:
%\begin{description}
%	\item[Locally supported impulse responses:] For all $x\in \Omega$, the support of $\impulseresponse_x$ is contained (or approximately contained) in a localized region.
%	\item[Non-negative impulse responses:] We have
%	\begin{equation*}
%	\impulseresponse_x(y) \ge 0
%	\end{equation*}
%	for all (or approximately all) $(y,x) \in \Omega \times \Omega$.
%%	\item[Matrix-free:] One cannot easily evaluate kernel entries $\Aker(y,x)$. Instead, access to $\Aop$ is available through ``operator actions.'' That is, we have a black-box computational procedure (which is typically expensive) through which we may evaluate the maps
%%	\begin{equation*}
%%	u \mapsto\Aop u \quad \text{and} \quad v \mapsto \Aop^Tv
%%	\end{equation*} 
%%	for arbitrary functions $u$ and $v$. After discretization (see Appendix \ref{app:discretized_operations}), $\Aop$ becomes a dense matrix, $\bm{A}$, that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means that one can perform matrix-vector products of $\bm{A}$ and $\bm{A}^T$ with arbitrary vectors, but one cannot easily\footnote{One could compute $\bm{v} = \bm{A}\bm{e}_j$, where $\bm{e}_j=(0,\dots,0,1,0,\dots,0)$ is the unit vector which has $j^\text{th}$ component equal to one and all other components equal to zero, then extract $\bm{A}_{ij} = \bm{v}_i$. However, this process is wasteful because one computes the entire vector $\bm{v}$, then discards all but one component.} access matrix entries $\bm{A}_{ij}$. In applications, applying $\bm{A}$ to a vector may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. 
%%	%For example, in reduced space approaches to PDE-constrained optimization and inverse problems, Hessian information is available only through the application of the Hessian to vectors, and one such Hessian application requires solving two PDEs \cite{HESSIANADJOINT}. 
%\end{description} 
%Once the hierarchical matrix approximation $\mathbf{A}_H$ is constructed, one may use H-matrix methods to perform further useful linear algebra operations involving $\mathbf{A}_H$, including matrix-matrix addition, matrix-matrix multiplication, matrix-vector products, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. WHY DO THESE OPERATIONS: KRYLOV SOLVERS, SAMPLING STRATEGIES, CAN'T DO WITH A.
%
%% approximate operators which possess the following properties:
%% 
%%We further limit our focus to operators that have
%%\textbf{Local support}
%%We say that $\Aop$ has local support if, for all $x\in \Omega$, the support of $\impulseresponse_x$ is contained (or approximately contained) in a small region. 
%%The smaller these supports are, the better our algorithm will perform.  Let $\delta_x$ denote the delta distribution (point source) centered at $x$. From \eqref{eq:impulse_response_delta_action}, we see that this local support property means that the response of $\Aop$ to a point source at $x$ is zero (or small) at points $y$ that are far from $x$.
%%\item[3) Local translation invariance:] We say that $\Aop$ is locally translation invariant if
%%\begin{equation}
%%	\label{eq:local_translation_invariance}
%%	\Aker(y+h, x+h) \approx \Aker(y,x)
%%\end{equation}
%%when $h$ is not too large. One can imagine the kernel entry $\Aker(y,x)$ as representing the strength of an interaction between a source at point $x$ and a target at point $y$. Local translation invariance means that the interaction strength remains roughly the same if the source and target points are both translated by the same vector. This is illustrated in Figure FIG.
%%\textbf{Non-negative impulse response:} We have
%%\begin{equation*}
%%\impulseresponse_x(y) \ge 0
%%\end{equation*}
%%for all $(y,x) \in \Omega \times \Omega$.

@@

The most popular matrix-free operator approximation methods are based on low rank approximation \cite{CCGOPAPERS}. This includes the Lanczos method, the randomized singular value decomposition \cite{HMTRANDOM}, and CUR approximation/skeletonization \cite{CUR,MAHONEY}. However, low rank approximation is not suitable for high rank operators, and many operators of practical interest are high rank. In the context of Hessians, methods based on low rank approximation suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator actions are required to form an accurate ``low rank'' approximation \cite{MYDISSERTATION}. 
%The ideal scenario from a scientific perspective (highly informative data) is therefore the worst case scenario from a computational perspective (large computational cost) \cite{MYDISSERTATION}.

%In general, product-convolution approximations take the form shown in \eqref{eq:product_convolution}, but $\convkernel_i$ and $w_i$ may be arbitrary functions.
Product-convolution (PC) approximations are approximations of an operator by weighted sums of convolution operators with spatially varying weights. A single convolution operator may be high rank, so PC approximations can (where applicable) approximate high rank operators efficiently. PC approximations have been used in a variety of fields going back several decades \cite{PRODCONVLIST}. Often PC methods are known as ``point spread function'' (PSF) methods, particularly in optics and image processing. There, the ``operator'' $\Aop$ is a physical process of distortion and blurring of light rays passing through an optical system such as a telescope or microscope. However, ``PSF'' is sometimes used as an umbrella term to refer to other approximation methods that focus on interpolating impulse responses or impulse response features.
%, including CUR-based low rank approximation \cite{NAGYPSFCUR}.
For background on PC and PSF methods, we recommend the following papers: \cite{PRODCONVGOOD}.

Our method is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses of $\Aop$ associated with a collection of points $x_i \in \Omega$ \cite{PRODCONVLIST}. A popular choice for PC methods in this class is to choose the $x_i$ to be nodes in a regular grid \cite{NAGY}.
%, and interpolate the convolution operators with piecewise linear interpolation \cite{NAGY}, splines \cite{PRODCONVSPLINES}, or radial basis functions \cite{PRODCONVRBF}. 
With a regular grid, a large number of $x_i$ are typically required to achieve an accurate approximation. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large \cite{PRODCONVMYPAPER}. But even with adaptive refinement, many $x_i$, and hence operator actions of $\Aop$, are required.

In this paper, 
%rather than using adaptive refinement, 
we reduce the computational cost by picking points $x_i$ in a way that allows many $\phi_{x_i}$ to be computed with each operator action of $\Aop$. This requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. We were inspired by techniques from resolution analysis in seismic imaging \cite{RESOLUTION}. In resolution analysis, $\Aop$ is the Hessian for a seismic inverse problem, and the width of $\phi_x$ is used to estimate of the minimum length scale on which features of the parameter can be inferred near the point $x$. If the width of $\phi_x$ is ten meters, then features of the parameter near $x$ can be accurately inferred from data if those features are larger than roughly ten meters in size. In \cite{RESOLUTION}, the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Sections \ref{sec:intromoments} and \ref{eq:mean_and_covariance_estimation}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing, but our method requires that $\Aop$ has a positive integral kernel, while random noise probing does not have this requirement.
%(and moreover, the seismic Hessian does not have a non-negative integral kernel).

Convolutions can be performed using the fast Fourier transform (FFT) \cite{PRODCONVFFT}, which allows PC approximations to be used as a fast surrogates for $\Aop$. Often it is necessary to perform further linear algebra operations involving $\Aop$, such as solving linear systems with $\Aop$ as the coefficient operator. Conventionally, one performs these linear algebra operations using Krylov methods, replacing $\Aop$ with its PC approximation \cite{PRODCONVKRYLOV}. The Krylov method is accelerated because performing matrix vector products with the product-convolution approximation is typically faster than performing operator actions with $\Aop$. However, there are several problems with this approach. First, the number of Krylov iterations can be large, particularly when $\Aop$ is high-rank. Second, complicated geometries are rarely discretized using regular grids, so using the FFT to perform convolutions requires transferring discretized functions back and forth between the irregular mesh and a regular grid. Third, the computational cost grows with the number of convolution kernels used. 
%If a large number of convolution kernels are used, this approach can be slow.
%Furthermore, in big-data inverse problems the required number of Krylov iterations is large \cite{MYDISSERTATION,OTHERS}. 
Thus, 
%after constructing the product-convolution approximation, 
it is desirable to convert the PC approximation to other matrix formats that are amenable to faster linear algebra operations. Wavelet compression methods have been used for this purpose \cite{PRODCONVWAVELETS}, because operators that are amenable to PC approximation are often sparse in the wavelet domain. Convolution-product approximations (transposes of PC approximations) are equivalent to low rank approximation of the pseudo-differential symbol of the operator, so one could perform linear algebra operations involving PC approximations using fast methods for pseudo-differential operator symbol calculus \cite{YINGDEMANET}. Other matrix-free operator approximation methods based on wavelets \cite{HERRMANN} and pseudo-differential operators \cite{DEMANET} do not use PC approximation. Here, we follow our previous work \cite{PRODCONVMYPAPER}, in which we converted PC approximations into H-matrix format.

Since our end goal is constructing an H-matrix, we are freed from strict adherence to the PC format. This allows us to modify the PC approach to address well-known deficiencies. PC approximations are based on an assumption of local translation invariance \cite{LOCATRANSLATIONINVARIANCE}, while our method is based on an assumption that we call ``normalized local mean displacement invariance,'' which is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). PC approximations have well-known issues near boundaries called boundary artifacts, which are caused by the boundary artificially ``cutting off'' impulse responses. 
%One must extend the impulse responses outside of the domain where they are undefined, but
%. Various extension techniques have been used, such as extension by zero, and reflection across the boundary, yet 
%no known extension technique is fully satisfactory \cite{BOUNDARYETENSION}. 
In our method we do not have an issue with boundary artifacts because we temporarily exclude impulse responses from the approximation procedure whenever the approximation procedure attempts to evaluate the impulse response at a location where it is undefined. In exchange for these benefits, our approximation cannot be written as a weighted sum of convolution operators, so we cannot apply our approximation to vectors via the FFT. This is fine because our approximation is an H-matrix, which can be applied to vectors even faster.

%MOMENTS PERMIT IMPROVED INTERPOLATION. IMPROVED INTERPOLATION REQUIRES H-MATRIX. H-MATRIX WITH IMPROVED INTERPOLATION ALSO SOLVES BOUNDARY ARTIFACT PROBLEM. ALSO DOESNT REQUIRE REGULAR GRID. ALSO SCALABLE TO LARGE NUMBER OF IMPULSE RESPONSES. FFT HAS FAST MATVEC BUT NOT OTHER OPTIONS.

Since we are building an H-matrix approximation of $\Aop$, it is natural to ask why we need to compute impulse responses and interpolate them. Why not build the H-matrix directly?
%Can we build the H-matrix directly? In short, no, not efficiently. 
The reason is that classical H-matrix construction methods \cite{HMATRIXCONSTRUCTION} require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. Our impulse response interpolation lets us quickly compute approximations to matrix entries, which allows us to use these classical H-matrix construction methods. There are newer matrix-free methods for H-matrix approximation based on the so-called ``peeling process'' \cite{LEXINGPEELINGPROCESS}, and these methods have been used to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{ILONAHMATRIX,NOEMIHMATRIX}. While these matrix-free H-matrix construction methods are asymptotically scalable in theory, in practice the required number of matrix vector products is unreasonably large. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods more feasible in the future.

%Product-convolution type approximations of Hessians have been used in a seismic inverse problem in \cite{GEORGSEISMICPRODCONV}, and in an advection-diffusion inverse problem in \cite{PRODCONVMYPAPER}. In \cite{DEMANETSEISMIC}, matrix probing \cite{MATRIXPROBING} is used to approximate the Hessian in a seismic inverse problem as the sum of simple pseudodifferential operators. Although it is not explicitly mentioned, the approximation in \cite{DEMANETSEISMIC} could be interpreted as a convolution-product approximation (a product-convolution approximation of the transpose of the operator).
%, which is like a product-convolution approximation, except the order of pointwise multiplications and convolutions is reversed in \eqref{eq:product_convolution}.

@@

The impulse response $\impulseresponse_x$ may be interpreted as a scaled probability distribution because of the non-negative integral kernel property. Let $\spatialvol:\Omega \rightarrow \mathbb{R}$,
\begin{equation*}
\spatialvol(x) := \int_\Omega \impulseresponse_x(y) dy
\end{equation*}
denote the spatially varying scaling factor,
%, let 
%\begin{equation}
%\label{eq:scaled_impulse_response}
%\widehat{\impulseresponse}_x := \impulseresponse_x / \spatialvol(x)
%\end{equation}
%denote the normalized version of $\impulseresponse_x$. 
and let $\mu:\Omega \rightarrow \mathbb{R}^d$, $\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d},$
\begin{equation*}
\spatialmean(x) := \mathbb{E}(\impulseresponse_x / V(x)), \qquad
\spatialcov(x) := \Var(\impulseresponse_x /V(x)),
\end{equation*}
to be the spatially varying mean and covariance of the normalized version of $\impulseresponse_x$, respectively. Imagine $\phi_x$ as a mound of dirt piled on a localized region in $\Omega$, with $\phi_x(y)$ being the height of the dirt mound at location $y$. The scalar $V(x)$ is the total volume of dirt in the mound, the vector $\mu(x)$ is the center of mass of the dirt mound, and the matrix $\Sigma(x)$ contains information about how wide the dirt mound is in each spatial direction.

>>>>

% REQUIRED
\begin{abstract}
	We present a fast matrix-free method for computing hierarchical matrix (H-matrix) approximations of operators that have locally supported non-negative integral kernels. Such operators arise, for example, as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained inverse problems. Fast H-matrix methods can add, multiply, invert, and factorize H-matrices, compute matrix-vector products, solve linear systems, and perform other useful matrix computations. Our method computes impulse responses of the operator at a collection of scattered points, then interpolates these known impulse responses to evaluate entries of the integral kernel at other points where the impulse response is not known. Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. 
	We apply the method to approximate Hessians in large-scale PDE-constrained inverse problems with highly informative data. Numerical results demonstrate that our method substantially outperforms existing state-of-the-art Hessian approximation methods which are based on low-rank approximation. We are able to form high quality approximations of high rank operators using only a small number of operator actions products.
\end{abstract}

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$) and let $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the following form:
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, $\Aop u \in L^2(\Omega)'$ is the linear functional that results from applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$. We seek to approximate operators $\Aop$ that are accessible \emph{matrix-free}. That is, we have a black-box computational procedure through which we may evaluate the maps
\begin{equation*}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv
\end{equation*} 
for arbitrary functions $u$ and $v$, but we cannot easily compute kernel entries $\Aker(y,x)$. In applications, evaluating these maps may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representation of $\Aop$, but cannot easily compute individual matrix entries.

Our method, described in Section \ref{sec:method}, belongs to a class of methods that approximate an operator by interpolating impulse responses of the operator associated with a scattered collection of points. The \emph{impulse response} of $\Aop$ associated with the point $x$ is the function $\impulseresponse_x:\Omega \rightarrow \mathbb{R}$, 
\begin{equation}
\label{eq:impulse_response_defn}
\impulseresponse_x(y) := \Aker(y, x),
\end{equation} 
which results from keeping $x$ fixed and viewing $\Phi(y,x)$ as a function of $y$. The function $\impulseresponse_x$ is called the impulse response because straightforward analysis shows that it may be expressed as follows:
\begin{equation}
\label{eq:impulse_response_delta_action}
\impulseresponse_x = \left( \Aop \delta_x^* \right)^*,
\end{equation}
where $\delta_x$ is the the delta distribution (i.e., point source, impulse) centered at $x$, $\Aop \delta_x^* \in L^2(\Omega)'$ is an abuse of notation that denotes the result of applying $\Aop$ to the distribution $\delta_x$, and $\left( \Aop \delta_x^* \right)^* \in L^2(\Omega)$ denotes the Riesz representation of $\Aop \delta_x^*$ with respect to the $L^2$ inner product.\footnote{Recall that the Riesz representative of a functional $\sigma \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\sigma^* \in L^2(\Omega)$ such that $\sigma(v) = \left(\sigma^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} Note that while the domain of $\Aop$ is defined as $L^2(\Omega)$, which does not contain $\delta_x$, if $\Aker$ is sufficiently regular then the action of $\Aop$ may be extended to distributions. This is described in Appendix \ref{app:distributions}.

@@

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch1.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch12.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch19.png}
	\end{subfigure}
	\caption{Batches of normalized impulse responses for the Stokes inverse problem Hessian. Left: 2nd batch. Mid: 13th batch. Right: 19th batch. Brightness scale differs between the subfigures. Black stars indicate the sample point locations at which the point source is applied. Dashed gray ellipses indicate the estimated support ellipsoids. The large circle is the boundary of $\Omega$. Early batches tend to have lots of ellipsoids near the boundary, as compared to later batches. This is because our greedy ellipsoid packing algorithm tries to pick sample points that are far away from sample points in previously chosen batches.}
	\label{fig:batches_intro}
\end{figure}

@@

@@

The most popular matrix-free operator approximation methods are based on low rank approximation \cite{CCGOPAPERS}. This includes the Lanczos method, the randomized singular value decomposition \cite{HMTRANDOM}, and CUR approximation/skeletonization \cite{CUR,MAHONEY}. However, low rank approximation is not suitable for high rank operators, and many operators of practical interest are high rank. In the context of Hessians, methods based on low rank approximation suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator actions are required to form an accurate ``low rank'' approximation \cite{MYDISSERTATION}.

Product-convolution (PC) approximations are approximations of an operator by weighted sums of convolution operators with spatially varying weights. A single convolution operator may be high rank, so PC approximations can (where applicable) approximate high rank operators efficiently. PC approximations have been used in a variety of fields going back several decades \cite{PRODCONVLIST}. Often PC methods are known as ``point spread function'' (PSF) methods, particularly in optics and image processing. There, the ``operator'' $\Aop$ is a physical process of distortion and blurring of light rays passing through an optical system such as a telescope or microscope. However, ``PSF'' is sometimes used as an umbrella term to refer to other approximation methods that focus on interpolating impulse responses or impulse response features. For background on PC and PSF methods, we recommend the following papers: \cite{PRODCONVGOOD}.

Our method is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses of $\Aop$ associated with a collection of points $x_i \in \Omega$ \cite{PRODCONVLIST}. A popular choice for PC methods in this class is to choose the $x_i$ to be nodes in a regular grid \cite{NAGY}. With a regular grid, a large number of $x_i$ are typically required to achieve an accurate approximation. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large \cite{PRODCONVMYPAPER}. But even with adaptive refinement, many $x_i$, and hence operator actions of $\Aop$, are required.

In this paper, we reduce the computational cost by picking points $x_i$ in a way that allows many $\phi_{x_i}$ to be computed with each operator action of $\Aop$. This requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. We were inspired by techniques from resolution analysis in seismic imaging \cite{RESOLUTION}. In resolution analysis, $\Aop$ is the Hessian for a seismic inverse problem, and the width of $\phi_x$ is used to estimate of the minimum length scale on which features of the parameter can be inferred near the point $x$. If the width of $\phi_x$ is ten meters, then features of the parameter near $x$ can be accurately inferred from data if those features are larger than roughly ten meters in size. In \cite{RESOLUTION}, the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Sections \ref{sec:intromoments} and \ref{eq:mean_and_covariance_estimation}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing, but our method requires that $\Aop$ has a positive integral kernel, while random noise probing does not have this requirement.

Convolutions can be performed using the fast Fourier transform (FFT) \cite{PRODCONVFFT}, which allows PC approximations to be used as a fast surrogates for $\Aop$. Often it is necessary to perform further linear algebra operations involving $\Aop$, such as solving linear systems with $\Aop$ as the coefficient operator. Conventionally, one performs these linear algebra operations using Krylov methods, replacing $\Aop$ with its PC approximation \cite{PRODCONVKRYLOV}. The Krylov method is accelerated because performing matrix vector products with the product-convolution approximation is typically faster than performing operator actions with $\Aop$. However, there are several problems with this approach. First, the number of Krylov iterations can be large, particularly when $\Aop$ is high-rank. Second, complicated geometries are rarely discretized using regular grids, so using the FFT to perform convolutions requires transferring discretized functions back and forth between the irregular mesh and a regular grid. Third, the computational cost grows with the number of convolution kernels used. Thus, it is desirable to convert the PC approximation to other matrix formats that are amenable to faster linear algebra operations. Wavelet compression methods have been used for this purpose \cite{PRODCONVWAVELETS}, because operators that are amenable to PC approximation are often sparse in the wavelet domain. Convolution-product approximations (transposes of PC approximations) are equivalent to low rank approximation of the pseudo-differential symbol of the operator, so one could perform linear algebra operations involving PC approximations using fast methods for pseudo-differential operator symbol calculus \cite{YINGDEMANET}. Other matrix-free operator approximation methods based on wavelets \cite{HERRMANN} and pseudo-differential operators \cite{DEMANET} do not use PC approximation. Here, we follow our previous work \cite{PRODCONVMYPAPER}, in which we converted PC approximations into H-matrix format.

Since our end goal is constructing an H-matrix, we are freed from strict adherence to the PC format. This allows us to modify the PC approach to address well-known deficiencies. PC approximations are based on an assumption of local translation invariance \cite{LOCATRANSLATIONINVARIANCE}, while our method is based on an assumption that we call ``normalized local mean displacement invariance,'' which is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). PC approximations have well-known issues near boundaries called boundary artifacts, which are caused by the boundary artificially ``cutting off'' impulse responses. In our method we do not have an issue with boundary artifacts because we temporarily exclude impulse responses from the approximation procedure whenever the approximation procedure attempts to evaluate the impulse response at a location where it is undefined. In exchange for these benefits, our approximation cannot be written as a weighted sum of convolution operators, so we cannot apply our approximation to vectors via the FFT. This is fine because our approximation is an H-matrix, which can be applied to vectors even faster.

Since we are building an H-matrix approximation of $\Aop$, it is natural to ask why we need to compute impulse responses and interpolate them. Why not build the H-matrix directly? The reason is that classical H-matrix construction methods \cite{HMATRIXCONSTRUCTION} require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. Our impulse response interpolation lets us quickly compute approximations to matrix entries, which allows us to use these classical H-matrix construction methods. There are newer matrix-free methods for H-matrix approximation based on the so-called ``peeling process'' \cite{LEXINGPEELINGPROCESS}, and these methods have been used to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{ILONAHMATRIX,NOEMIHMATRIX}. While these matrix-free H-matrix construction methods are asymptotically scalable in theory, in practice the required number of matrix vector products is unreasonably large. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods more feasible in the future.

@@

The impulse response $\impulseresponse_x$ may be interpreted as a scaled probability distribution because of the non-negative integral kernel property. Let $\spatialvol:\Omega \rightarrow \mathbb{R}$,
\begin{equation*}
\spatialvol(x) := \int_\Omega \impulseresponse_x(y) dy
\end{equation*}
denote the spatially varying scaling factor, and let $\mu:\Omega \rightarrow \mathbb{R}^d$, $\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d},$
\begin{equation*}
\spatialmean(x) := \mathbb{E}(\impulseresponse_x / V(x)), \qquad
\spatialcov(x) := \Var(\impulseresponse_x /V(x)),
\end{equation*}
to be the spatially varying mean and covariance of the normalized version of $\impulseresponse_x$, respectively. Imagine $\phi_x$ as a mound of dirt piled on a localized region in $\Omega$, with $\phi_x(y)$ being the height of the dirt mound at location $y$. The scalar $V(x)$ is the total volume of dirt in the mound, the vector $\mu(x)$ is the center of mass of the dirt mound, and the matrix $\Sigma(x)$ contains information about how wide the dirt mound is in each spatial direction.
